---
title: "MINI SAM DISTATIS ANALYSIS"
author:
  - Soudeh Khoubrouy
  - Ryan C. Yeung
  - Carina L. Fan
  - Divya Mistry
  - Ju-Chi Yu
  - Aida Eslami
  - Hervé Abdi
  - Brian Levine
date: last-modified 04/08/2026
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

# Overview

This report details the application of the **DiSTATIS** method to a series of Mini SAM (Survey of Autobiographical Memory) datasets. Mini SAM only contains 10 questions of the 26 questions of SAM. DiSTATIS [Abdi et al., 2012](https://doi.org/10.1002/wics.198) combines information from multiple studies that measure the same set of variables. In the first step of the analysis an item-by-item correlation matrix is computed. Each correlation matrix is then double centered and normalized.
Next, DISTATIS creates the compromise correlation matrix that represents the best overall summary of the cross-product matrices by computing a weighted average of the previous double centered and normalized matrices. In this weighted combination, higher weights are assigned to studies that more closely resemble the others.
Eigen decomposition of the compromise matrix identifies the main directions, or components, along which the variables vary the most. These components define a space that shows how the variables relate to one another. Variables that behave similarly appear closer together (with smaller angle) in this space, while variables that differ appear farther apart (with larger angle). 
To explore each study’s partial factorial structure, we projected each study onto the compromise space. These partial structures show how each study positions the SAM variables relative to the global structure.  

The analysis focuses on:

1. **RV Analysis**: Assessing the similarity between different datasets.

2. **Compromise Analysis**: Identifying common patterns across all participants/studies and visualizing how specific datasets deviate from the common pattern.

# Reproducibility Notes

This Quarto file assumes a project structure similar to:

```text
project/
├── data/
│   ├── dataSAM.rda
│   ├── dataMiniSAM.rda
│   └── dataAgingSAM.rda
├── figures/
│   ├── SAM/
│   ├── miniSAM/
│   └── agingSAM/
├── functions/
│   ├── runMeFirst.R
│   ├── prepData.R
│   ├── plotPartialFactor.R
│   ├── plotFactor.R
│   ├── myCreatePartialFactorScoresMap.R
│   ├── myCreateFactorMap.R
│   └── myCorPlot.R
└── quartos/
│   ├── distatisAgingSAM.qmd
│   ├── distatisMiniSAM.qmd
│   └── distatisSAM.qmd  
└── README.md
```

If your folder structure is different, update the relative paths in the setup
section.

# Setup

## Clear Workspace and Graphics Devices

```{r}
rm(list = ls())
graphics.off()
```

## Install Required Packages

```{r}
source("../functions/runMeFirst.R")
```

## Load Required Packages

```{r}
library(prettyGraphs)
library(PTCA4CATA)
library(data4PCCAR)
library(DistatisR)
library(gplots)
library(ggrepel)
library(ggplot2)
library(corrplot)
library(gridExtra)
library(ggplotify)
library(grid)
library(rgl)
```
Note: If any of the required packages are not installed on your system, you can install them using:

```{r eval=FALSE}
install.packages("package_name")
```

## Source Project-Specific Functions

```{r}
source("../functions/prepData.R")
source("../functions/plotPartialFactor.R")
source("../functions/plotFactor.R")
source("../functions/myCreatePartialFactorScoresMap.R")
source("../functions/myCreateFactorMap.R")
source("../functions/myCorPlot.R")
```

# Load Data

We processed the .xlsx data file for each dataset and constructed a cross-product matrix ($X^T X$) after centering and normalizing the data. We then concatenated all the cross-product matrices in a cube.
The number of Mini SAM variables is 10, and in our case, we have 28 datasets. As a result, the Cube has dimensions 10x10x28, where the first two dimensions correspond to the SAM variables and the third dimension represents the different datasets.
To ensure data confidentiality, we provide the Cube as an .rda file instead of sharing the original datasets. Moreover, the datasets will be shown with numbers 1 through 28 instead of their actual names.

```{r}
dataCubeName <- "dataMiniSAM.rda"
load(file.path("../data", dataCubeName))
```

## Optional: Rebuild the Cube from Original Excel Files

If you have access to your original datasets and want to rebuild the cube
instead of loading `dataSAM.rda`, you can use the code below.

```{r eval=FALSE}
# Make sure all the excel files are under the data folder.
files <- list.files(
  path = "../data/",
  pattern = ".xlsx",
  full.names = TRUE
)

files <- files[!grepl("^~\$", basename(files))]
filesNames <- basename(files)
datasetSize <- numeric(length(files))
numSamVar <- 26
numSamDatasets <- length(files)
Cube <- array(0, c(numSamVar, numSamVar, numSamDatasets))

# Read Excel file
for (j in seq_along(files)) {
  data <- read.df.excel(path = files[j], sheet = "data")
  dataRaw <- as.data.frame(data$df.data)
  # Double center and normalize
  dataTData <- prepData(dataRaw, center = TRUE, scale = FALSE, norm = TRUE)
  Cube[, , j] <- dataTData
  datasetSize[j] <- nrow(data$df.data)
}

filesNames[order(datasetSize)] <- seq_along(files)
dimnames(Cube) <- list(rownames(dataTData), rownames(dataTData), filesNames)
```

# Define Colors and Labels

## Colors for SAM Variables

Below we use the color coding used in the original paper of SAM.

```{r}
numSamVar <- dim(Cube)[1]
namSamVar <- dimnames(Cube)[[1]]
filesNames <- dimnames(Cube)[[3]]
color4SAM <- character(numSamVar)

colorCode <- c("steelblue2", "darkolivegreen3", "tomato3", "purple4")

for (i in seq_len(numSamVar)) {
  if (grepl("e", namSamVar[i])) {
    color4SAM[i] <- colorCode[1]
  } else if (grepl("f", namSamVar[i])) {
    color4SAM[i] <- colorCode[2]
  } else if (grepl("m", namSamVar[i])) {
    color4SAM[i] <- colorCode[3]
  } else {
    color4SAM[i] <- colorCode[4]
  }
}
```

## Colors for Datasets

In the original workflow, dataset colors were based on dataset size (i.e., staleblue4 for size<100, chartreuse4 for 100=<size<300, yellow3 for size 300=<size<600, darkorang for 600=<size<1000, and red3 for larger datasets) . Because
this Quarto version starts from a saved cube, the final color vector is
specified directly.

```{r}

color4Datasets <- c(
  "slateblue4",  "slateblue4",  "chartreuse4",
  "red3",        "slateblue4",  "chartreuse4",
  "slateblue4",  "yellow3",     "darkorange",
  "red3",        "yellow3",     "yellow3",
  "darkorange",  "slateblue4",  "chartreuse4",
  "chartreuse4", "chartreuse4", "red3",
  "chartreuse4", "yellow3",     "chartreuse4",
  "yellow3",     "chartreuse4", "chartreuse4",
  "darkorange",  "chartreuse4", "darkorange",
  "slateblue4"
)
```

If you have access to your original datasets and dataset sizes, you can instead
generate these colors automatically using the following code:

```{r eval=FALSE}
# Required variables (files and color4Datasets) must be found in section 4.1.
color4Datasets <- character(length(files))
for (j in seq_along(files)) {
  if (datasetSize[j] < 100) {
    color4Datasets[j] <- "slateblue4"
  } else if (datasetSize[j] < 300) {
    color4Datasets[j] <- "chartreuse4"
  } else if (datasetSize[j] < 600) {
    color4Datasets[j] <- "yellow3"
  } else if (datasetSize[j] < 1000) {
    color4Datasets[j] <- "darkorange"
  } else {
    color4Datasets[j] <- "red3"
  }
}
```

# Run DiSTATIS

Below we use distatis function from DistatisR package to analyze the data. The analysis keeps the first 6 factors and uses MFA normalization with double
centering.

```{r}
nFact2Keep <- 6

resDistatis <- distatis(
  Cube,
  Distance = FALSE,
  Norm = "MFA",
  double_centering = TRUE,
  RV = TRUE,
  nfact2keep = nFact2Keep
)
```

## Align the Signs of Loadings and Scores

The sign of factor scores is arbitrary. The following code flips signs so that
the largest absolute loading on each dimension is positive.

```{r}
for (i in seq_len(ncol(resDistatis$res4Cmat$G))) {
  if (resDistatis$res4Cmat$G[which.max(abs(resDistatis$res4Cmat$G[, i])), i] < 0) {
    resDistatis$res4Cmat$G[, i] <- -resDistatis$res4Cmat$G[, i]
  }
}

for (i in seq_len(ncol(resDistatis$res4Splus$F))) {
  if (resDistatis$res4Splus$F[which.max(abs(resDistatis$res4Splus$F[, i])), i] < 0) {
    resDistatis$res4Splus$F[, i] <- -resDistatis$res4Splus$F[, i]
    resDistatis$res4Splus$PartialF[, i, ] <- -resDistatis$res4Splus$PartialF[, i, ]
  }
}
```

# RV Analysis

## RV Matrix Heatmap

Here is the correlation heatmap of the RV matrix.
```{r}
myCorPlot(
  data = resDistatis$res4Cmat$C,
  order = "FPC",
  full.type = FALSE,
  title = NULL,
  name.color = color4Datasets,
  is.corr = TRUE
)
```

## Dataset Weights (Alphas)

Below we see the weights used for the datasets. These weights have been found from RV matrix such that higher weights are assigned to studies that more closely resemble the others.

```{r}
barplot(
  resDistatis$res4Cmat$alpha,
  ylim = NULL,
  col = color4Datasets,
  ylab = "Weights for Datasets",
  cex.names = 0.7,
  cex.axis = 0.9,
  las = 2
)
```

## RV Scree Plot

Here is the scree plot of the RV matrix:

```{r}
PlotScree(
  ev = resDistatis$res4Cmat$eigValues,
  plotKaiser = TRUE,
  color4Kaiser = ggplot2::alpha("darkorchid4", 0.5),
  lwd4Kaiser = 2,
  title = NULL,
  xlab = "Components"
)
```

## RV Loading Map

Here is the loading map for the RV matrix:

```{r}
rvLoadingsList <- plotFactor(
  Fj = resDistatis$res4Cmat$G,
  lambda = resDistatis$res4Cmat$eigValues,
  tau = resDistatis$res4Cmat$tau,
  x = 1,
  y = 2,
  title = NULL,
  subtitle = NULL,
  points.color = color4Datasets,
  constraints.factor = 1.15
)

print(rvLoadingsList[[1]])
```

# Compromise Analysis

## Compromise Matrix Heatmap

Here is the heatmap for the Compromise matrix values:

```{r}
myCorPlot(
  data = resDistatis$res4Splus$Splus,
  full.type = FALSE,
  title = NULL,
  name.color = color4SAM,
  is.corr = FALSE
)
```

## Compromise Scree Plot

Here is the scree plot for the Compromise matrix:

```{r}
PlotScree(
  ev = resDistatis$res4Splus$eigValues,
  plotKaiser = TRUE,
  color4Kaiser = ggplot2::alpha("darkorchid4", 0.5),
  lwd4Kaiser = 2,
  title = NULL,
  xlab = "Components"
)
```

## Global and Partial Factor Scores with Confidence Intervals

This section generates compromise loading (Global factor score) maps, partial factor score maps, and
confidence interval ellipses based on bootstrap partial factor scores.

```{r}
comp <- list()

for (k in 1:3) {
  x <- 2 * k - 1
  y <- 2 * k

  # Compromise loadings
  comp[[k]] <- plotFactor(
    Fj = resDistatis$res4Splus$F,
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title = NULL,
    subtitle = NULL,
    points.color = color4SAM,
    constraints.factor = 1.2
  )

  print(comp[[k]][[1]])

  # Partial loadings
  partials <- plotPartialFactor(
    Fj = resDistatis$res4Splus$F,
    partialF = resDistatis$res4Splus$PartialF,
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title.datasets = NULL,
    subtitle.datasets = NULL,
    title.items = NULL,
    subtitle.items = NULL,
    names4Partials = filesNames,
    color4Blocks = color4Datasets,
    color4Items = color4SAM,
    points.color = color4SAM,
    minx = -2.8 * abs(min(resDistatis$res4Splus$F[, x])),
    maxx =  2.0 * abs(max(resDistatis$res4Splus$F[, x])),
    miny = -3.0 * abs(min(resDistatis$res4Splus$F[, y])),
    maxy =  2.5 * abs(max(resDistatis$res4Splus$F[, y]))
  )

  # Partial loadings colored by datasets
  print(partials[[1]])
  # Partial loadings colored by SAM variables
  print(partials[[2]])

  # Confidence intervals

  bootF <- BootFactorScores(resDistatis$res4Splus$PartialF)
  
  pfsElli <- MakeCIEllipses(
    data = bootF[, c(x, y), ],
    names.of.factors = c(
      paste0("Factor ", x),
      paste0("Factor ", y)
    ),
    col = color4SAM,
    line.size = 0.8
  )
  
  # Compromise loadings and confidence intervals
  print(comp[[k]][[1]] + pfsElli)

}
```
## Contribution Plots

### Dataset Contributions

Here is the contribution made by each dataset to each of the components:

```{r}
# Dataset contributions
signedCtr <- resDistatis$res4Cmat$ctr * sign(resDistatis$res4Cmat$G)
ctr <- list()

for (k in 1:3) {
  ctr[[k]] <- PrettyBarPlot2(
    signedCtr[, k],
    threshold = 1 / nrow(signedCtr),
    font.size = 2,
    color4bar = gplots::col2hex(color4Datasets),
    ylab = "Contributions",
    ylim = c(
      -1.2 * abs(min(signedCtr[, k])),
      1.2 * max(signedCtr[, k])
    )
  ) +
    ggtitle(paste0("Component ", k)) +
    theme(
      plot.title = element_text(size = 9),
      axis.title.y = element_text(size = 9)
    )
}

grid.arrange(
  as.grob(ctr[[1]]),
  as.grob(ctr[[2]]),
  as.grob(ctr[[3]]),
  ncol = 1,
  nrow = 3,
  top = textGrob("", gp = gpar(fontsize = 18, font = 3))
)
```

### SAM Variable Contributions

Here is the contribution made by SAM variables to each of the components:

```{r}
# SAM variable contributions
signedSamCtr <- resDistatis$res4Splus$ctr * sign(resDistatis$res4Splus$F)
samCtr <- list()

for (k in 1:6) {
  samCtr[[k]] <- PrettyBarPlot2(
    signedSamCtr[, k],
    threshold = 1 / nrow(signedSamCtr),
    font.size = 2,
    color4bar = gplots::col2hex(color4SAM),
    ylab = "Contributions",
    ylim = c(
      -1.2 * abs(min(signedSamCtr[, k])),
      1.2 * max(signedSamCtr[, k])
    )
  ) +
    ggtitle(paste0("Component ", k)) +
    theme(
      plot.title = element_text(size = 9),
      axis.title.y = element_text(size = 9)
    )
}

grid.arrange(
  as.grob(samCtr[[1]]),
  as.grob(samCtr[[2]]),
  as.grob(samCtr[[3]]),
  as.grob(samCtr[[4]]),
  as.grob(samCtr[[5]]),
  as.grob(samCtr[[6]]),
  ncol = 2,
  nrow = 3,
  top = textGrob("", gp = gpar(fontsize = 18, font = 3))
)
```

# Varimax Rotation Effect on Compromise Loadings

Here, we show the Varimax rotation effect.

```{r}
# Apply Varimax rotation
variCompromise <- varimax(resDistatis$res4Splus$F[, 1:4], normalize = FALSE)
fjVarimaxed <- resDistatis$res4Splus$F[, 1:4] %*% variCompromise$rotmat

compVarimaxed <- list()

for (k in 1:2) {
  x <- 2 * k - 1
  y <- 2 * k

  compVarimaxed[[k]] <- plotFactor(
    Fj = fjVarimaxed,
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title = NULL,
    subtitle = NULL,
    points.color = color4SAM,
    constraints.factor = 1.5
  )
    print(compVarimaxed[[k]][[1]])
}
```


