---
title: "SAM DISTATIS ANALYSIS FOR AGE EFFECT"
author:
  - Soudeh Khoubrouy
  - Ryan C. Yeung
  - Carina L. Fan
  - Divya Mistry
  - Ju-Chi Yu
  - Aida Eslami
  - Hervé Abdi
  - Brian Levine
date: last-modified 04/08/2026
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This report details the application of the **DiSTATIS** method to a series of SAM (Survey of Autobiographical Memory) datasets and shows the age effect on SAM. DiSTATIS [Abdi et al., 2012](https://doi.org/10.1002/wics.198) combines information from multiple studies that measure the same set of variables. In the first step of the analysis an item-by-item correlation matrix is computed. Each correlation matrix is then double centered and normalized.
Next, DISTATIS creates the compromise correlation matrix that represents the best overall summary of the cross-product matrices by computing a weighted average of the previous double centered and normalized matrices. In this weighted combination, higher weights are assigned to studies that more closely resemble the others.
Eigen decomposition of the compromise matrix identifies the main directions, or components, along which the variables vary the most. These components define a space that shows how the variables relate to one another. Variables that behave similarly appear closer together (with smaller angle) in this space, while variables that differ appear farther apart (with larger angle). 
To explore each study’s partial factorial structure, we projected each study onto the compromise space. These partial structures show how each study positions the SAM variables relative to the global structure.  

The analysis focuses on:

1. **RV Analysis**: Assessing the similarity between different datasets.

2. **Compromise Analysis**: Identifying common patterns across all participants/studies and visualizing how specific datasets deviate from the common pattern.

# Reproducibility Notes

This Quarto file assumes a project structure similar to:

```text
project/
├── data/
│   ├── dataSAM.rda
│   ├── dataMiniSAM.rda
│   └── dataAgingSAM.rda
├── figures/
│   ├── SAM/
│   ├── miniSAM/
│   └── agingSAM/
├── functions/
│   ├── prepData.R
│   ├── plotPartialFactor.R
│   ├── plotFactor.R
│   ├── myCreatePartialFactorScoresMap.R
│   ├── myCreateFactorMap.R
│   └── myCorPlot.R
└── quartos/
│   ├── distatisAgingSAM.qmd
│   ├── distatisMiniSAM.qmd
│   └── distatisSAM.qmd  
└── README.md
```

If your folder structure is different, update the relative paths in the setup
section.

# Setup

## Clear Workspace and Graphics Devices

```{r}
rm(list = ls())
graphics.off()
```

## Install Required Packages

```{r}
source("../functions/runMeFirst.R")
```

## Load Required Packages

```{r}
library(prettyGraphs)
library(PTCA4CATA)
library(data4PCCAR)
library(DistatisR)
library(gplots)
library(ggrepel)
library(ggplot2)
library(corrplot)
library(gridExtra)
library(ggplotify)
library(grid)
library(rgl)
```
Note: If any of the required packages are not installed on your system, you can install them using:

```{r eval=FALSE}
install.packages("package_name")
```

## Source Project-Specific Functions

```{r}
source("../functions/prepData.R")
source("../functions/plotPartialFactor.R")
source("../functions/plotFactor.R")
source("../functions/myCreatePartialFactorScoresMap.R")
source("../functions/myCreateFactorMap.R")
source("../functions/myCorPlot.R")
```

# Load Data

We processed the .xlsx data file for each dataset and constructed a cross-product matrix ($X^T X$) after centering and normalizing the data. We then concatenated all the cross-product matrices of young datasets in one cube, and old datasets in another.
The number of SAM variables is 26, and in our case, we have 5 datasets with young participants and 6 datasets with old participants. Therefore, we have two cubes, one with dimensions 26x26x5 and one with dimensions 26x26x6, where the first two dimensions correspond to the SAM variables and the third dimension represents the different datasets.
To ensure data confidentiality, we provide the cubes as an .rda file instead of sharing the original datasets. 

```{r}
dataCubeName <- "dataAgingSAM.rda"
load(file.path("../data", dataCubeName))
```

# Define Colors and Labels

## Colors for SAM Variables

Below we use the color coding used in the original paper of SAM.

```{r}
numSamVar <- dim(CubeYoung)[1]
namSamVar <- dimnames(CubeYoung)[[1]]
color4SAM <- character(numSamVar)

colorCode <- c("steelblue2", "darkolivegreen3", "tomato3", "purple4")

for (i in seq_len(numSamVar)) {
  if (grepl("e", namSamVar[i])) {
    color4SAM[i] <- colorCode[1]
  } else if (grepl("f", namSamVar[i])) {
    color4SAM[i] <- colorCode[2]
  } else if (grepl("m", namSamVar[i])) {
    color4SAM[i] <- colorCode[3]
  } else {
    color4SAM[i] <- colorCode[4]
  }
}
```

## Colors for Datasets

```{r}
color4Datasets <- c("chartreuse4", "red3")
color4Alphas <- color4Datasets
```

# Run DiSTATIS on Young Datasets

Below we use distatis function from DistatisR package to analyze the data. The analysis keeps the first 6 factors and uses MFA normalization with double centering.
We first apply DiSTATIS on Young datasets.

```{r}
nFact2Keep <- 6

# Run DISTATIS on Young Data Sets ----
# Use MFA norm and double centering
resDistatis <- distatis(
  CubeYoung,
  Distance = FALSE,
  Norm = "MFA",
  double_centering = TRUE,
  RV = TRUE,
  nfact2keep = nFact2Keep
)

youngCompromise <- resDistatis$res4Splus$Splus
```

# Run DiSTATIS on Old Datasets

We then apply DiSTATIS on Old datasets.

```{r}
# Run DISTATIS on Old Data Sets ----
# Use MFA norm and double centering
resDistatis <- distatis(
  CubeOld,
  Distance = FALSE,
  Norm = "MFA",
  double_centering = TRUE,
  RV = TRUE,
  nfact2keep = nFact2Keep
)

oldCompromise <- resDistatis$res4Splus$Splus
```

# Create Final Cube

We create a cube of the young and Old Compromise matrices. This cube has 26x26x2 dimension.

```{r}
# Create Cube of Compromise Matrices for DISTATIS ----
numSamVar <- 26
CubeOfCompromises <- array(0, c(numSamVar, numSamVar, 2))

youngDataTData <- prepData(youngCompromise, center = TRUE, scale = FALSE, norm = TRUE)
CubeOfCompromises[, , 1] <- youngDataTData

oldDataTData <- prepData(oldCompromise, center = TRUE, scale = FALSE, norm = TRUE)
CubeOfCompromises[, , 2] <- oldDataTData

dimnames(CubeOfCompromises) <- list(
  rownames(youngDataTData),
  rownames(youngDataTData),
  c("Young", "Old")
)
```

# Run Final DiSTATIS

We apply DiSTATIS on the new cube.

```{r}
nFact2Keep <- 4

resDistatis <- distatis(
  CubeOfCompromises,
  Distance = FALSE,
  Norm = "MFA",
  double_centering = TRUE,
  RV = TRUE,
  nfact2keep = nFact2Keep
)
```

## Align the Signs of Loadings and Scores

The sign of factor scores is arbitrary. The following code flips signs so that
the largest absolute loading on each dimension is positive.

```{r}
for (i in seq_len(ncol(resDistatis$res4Cmat$G))) {
  if (resDistatis$res4Cmat$G[which.max(abs(resDistatis$res4Cmat$G[, i])), i] < 0) {
    resDistatis$res4Cmat$G[, i] <- -resDistatis$res4Cmat$G[, i]
  }
}

for (i in seq_len(ncol(resDistatis$res4Splus$F))) {
  if (resDistatis$res4Splus$F[which.max(abs(resDistatis$res4Splus$F[, i])), i] < 0) {
    resDistatis$res4Splus$F[, i] <- -resDistatis$res4Splus$F[, i]
    resDistatis$res4Splus$PartialF[, i, ] <- -resDistatis$res4Splus$PartialF[, i, ]
  }
}
```

# RV Analysis

## Dataset Weights (Alphas)

Below we see the weights used for the datasets. These weights have been found from RV matrix such that higher weights are assigned to studies that more closely resemble the others. Because there are only two main datasets, alpha will be the same for both.

```{r}
barplot(
  resDistatis$res4Cmat$alpha,
  ylim = NULL,
  col = color4Datasets,
  ylab = "Weights for Datasets",
  cex.names = 0.7,
  cex.axis = 0.9,
  las = 2
)
```

## RV Scree Plot

Here is the scree plot of the RV matrix:

```{r}
PlotScree(
  ev = resDistatis$res4Cmat$eigValues,
  plotKaiser = TRUE,
  color4Kaiser = ggplot2::alpha("darkorchid4", 0.5),
  lwd4Kaiser = 2,
  title = NULL,
  xlab = "Components"
)
```

## RV Loading Map

Here is the loading map for the RV matrix:

```{r}
rvLoadingsList <- plotFactor(
  Fj = resDistatis$res4Cmat$G,
  lambda = resDistatis$res4Cmat$eigValues,
  tau = resDistatis$res4Cmat$tau,
  x = 1,
  y = 2,
  title = NULL,
  subtitle = NULL,
  points.color = color4Datasets,
)

print(rvLoadingsList[[1]])
```

# Compromise Analysis

## Compromise Matrix Heatmap

Here is the heatmap for the Compromise matrix values:

```{r}
myCorPlot(
  data = resDistatis$res4Splus$Splus,
  full.type = FALSE,
  title = NULL,
  name.color = color4SAM,
  is.corr = FALSE
)
```

## Compromise Scree Plot

Here is the scree plot for the Compromise matrix:

```{r}
PlotScree(
  ev = resDistatis$res4Splus$eigValues,
  plotKaiser = TRUE,
  color4Kaiser = ggplot2::alpha("darkorchid4", 0.5),
  lwd4Kaiser = 2,
  title = NULL,
  xlab = "Components"
)
```

## Global and Partial Factor Scores with Confidence Intervals

This section generates compromise loading (Global factor score) maps and partial factor score maps.
Since there are only two main datasets (Young and Old) confidence interval ellipses will be lines and are not shown here.

```{r}
comp <- list()
youngPartials <- list()
oldPartials <- list()

for (k in 1:2) {
  x <- 2 * k - 1
  y <- 2 * k
  
  # Compromise loadings
  comp[[k]] <- plotFactor(
    Fj = resDistatis$res4Splus$F,
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title = NULL,
    subtitle = NULL,
    points.color = color4SAM,
    constraints.factor = 1.5
  )
  
  print(comp[[k]][[1]])
  
  # Partial loadings
  partials <- plotPartialFactor(
    Fj = resDistatis$res4Splus$F,
    partialF = resDistatis$res4Splus$PartialF,
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title.datasets = NULL,
    subtitle.datasets = NULL,
    title.items = NULL,
    subtitle.items = NULL,
    names4Partials = c("Y", "O"),
    color4Blocks = color4Datasets,
    color4Items = color4SAM,
    cex.point = 1.5,
    cex.text = 2.5,
    points.color = color4SAM,
    points.alpha = 0.2,
    labels.alpha = 0.6,
    axes.alpha = 0.6,
    minx = -1.5 * abs(min(resDistatis$res4Splus$F[, x])),
    maxx =  1.5 * abs(max(resDistatis$res4Splus$F[, x])),
    miny = -1.5 * abs(min(resDistatis$res4Splus$F[, y])),
    maxy =  1.5 * abs(max(resDistatis$res4Splus$F[, y]))
  )
  
  # Partial loadings colored by datasets
  print(partials[[1]])
  # Partial loadings colored by SAM variables
  print(partials[[2]])
  
  # Young partial loadings only
  youngPartials[[k]] <- plotFactor(
    Fj = resDistatis$res4Splus$PartialF[, , 1],
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title = "Young Partial Loadings",
    subtitle = NULL,
    points.color = color4SAM,
    constraints = comp[[k]][[2]]
  )
  
  print(youngPartials[[k]][[1]])
  
  # Old partial loadings only 
  oldPartials[[k]] <- plotFactor(
    Fj = resDistatis$res4Splus$PartialF[, , 2],
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title = "Old Partial Loadings",
    subtitle = NULL,
    points.color = color4SAM,
    constraints = comp[[k]][[2]]
  )
  
  print(oldPartials[[k]][[1]])
  
}
```

## Contribution Plots

### Dataset Contributions

Here is the contribution made by each dataset to each of the components:

```{r}
# Dataset contributions
signedCtr <- resDistatis$res4Cmat$ctr * sign(resDistatis$res4Cmat$G)
ctr <- list()

for (k in 1:2) {
  ctr[[k]] <- PrettyBarPlot2(
    signedCtr[, k],
    threshold = 1 / nrow(signedCtr),
    font.size = 4,
    color4bar = gplots::col2hex(color4Datasets),
    ylab = "Contributions",
    ylim = c(
      -1.2 * abs(min(signedCtr[, k])),
      1.2 * max(signedCtr[, k])
    )
  ) +
    ggtitle(paste0("Component ", k)) +
    theme(
      plot.title = element_text(size = 9),
      axis.title.y = element_text(size = 9)
    )
}

grid.arrange(
  as.grob(ctr[[1]]),
  as.grob(ctr[[2]]),
  ncol = 1,
  nrow = 2,
  top = textGrob("Signed Contributions", gp = gpar(fontsize = 18, font = 3))
)
```

### SAM Variable Contributions

Here is the contribution made by SAM variables to each of the components:

```{r}
# SAM variable contributions
signedSamCtr <- resDistatis$res4Splus$ctr * sign(resDistatis$res4Splus$F)
samCtr <- list()

for (k in 1:4) {
  samCtr[[k]] <- PrettyBarPlot2(
    signedSamCtr[, k],
    threshold = 1 / nrow(signedSamCtr),
    font.size = 4,
    color4bar = gplots::col2hex(color4SAM),
    ylab = "Contributions",
    ylim = c(
      -1.2 * abs(min(signedSamCtr[, k])),
      1.2 * max(signedSamCtr[, k])
    )
  ) +
    ggtitle(paste0("Component ", k)) +
    theme(
      plot.title = element_text(size = 9),
      axis.title.y = element_text(size = 9)
    )
}

grid.arrange(
  as.grob(samCtr[[1]]),
  as.grob(samCtr[[2]]),
  as.grob(samCtr[[3]]),
  as.grob(samCtr[[4]]),
  ncol = 2,
  nrow = 3,
  top = textGrob("", gp = gpar(fontsize = 18, font = 3))
)
```

# Varimax Rotation Effect on Compromise Loadings

Here, we show the Varimax rotation effect.

```{r}
# Apply Varimax rotation
variCompromise <- varimax(resDistatis$res4Splus$F[, 1:4], normalize = FALSE)
fjVarimaxed <- resDistatis$res4Splus$F[, 1:4] %*% variCompromise$rotmat

compVarimaxed <- list()

for (k in 1:2) {
  x <- 2 * k - 1
  y <- 2 * k

  compVarimaxed[[k]] <- plotFactor(
    Fj = fjVarimaxed,
    lambda = resDistatis$res4Splus$eigValues,
    tau = resDistatis$res4Splus$tau,
    x = x,
    y = y,
    title = NULL,
    subtitle = NULL,
    points.color = color4SAM,
    constraints.factor = 1.5
  )
    print(compVarimaxed[[k]][[1]])
}
```


